---
title: "SpurAfrika2018"
always_allow_html: yes
output:
  pdf_document: default
  html_document:
    keep_md: yes
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 12, fig.height = 8) # larger plot size
```

# Data import and processing

``` {r results='hide', message=FALSE, warning=FALSE}
# start necessary libraries

library(airtabler)
# airtabler required to read airtable database
# airtabler is installed by 'devtools::install_github("bergant/airtabler")'
# devtools can fail if the 'unzip' settings has not been set properly
# it can be set by 'options(unzip="unzip")'

library(tidyverse)
library(lubridate)
# data management
# time management

library(knitr)
# reproducible research and markdown library

# plotting libraries
library(ggplot2)
library(plotly)
library(webshot)
```

``` {r}
# read in data
ClinicData <- 
  airtable(
    base = 'appXwAbxa7r6bkKnO',
    tables = c('SpurAfrika2018','Schools','FindingCodes','ManagementCodes')
  )

RawPatientData <- as.tibble(ClinicData$SpurAfrika2018$select_all())
RawSchoolData <- as.tibble(ClinicData$Schools$select_all())
RawFindingCodes <- as.tibble(ClinicData$FindingCodes$select_all())
RawManagementCodes <- as.tibble(ClinicData$ManagementCodes$select_all())

# turn variable names into valid R names e.g. remove spaces
names(RawPatientData) <- make.names(names(RawPatientData))
```

## Adjust and re-calculate data

Attach variable **SchoolName** to each student.
Variable **School.Name** contains the school's database ID

Recalculate **Age.in.years** with **Data.collection.date** and **Date.of.birth**

``` {r}
# find actual ctual name (instead of the ID)
RawPatientData <- RawPatientData %>%
  mutate(SchoolName = RawSchoolData$SchoolName[match(School.Name,RawSchoolData$id)])

# re-calculate age in years
RawPatientData <- RawPatientData %>%
  mutate(Age.in.years = as.numeric(as.Date(Data.collection.date)-as.Date(Date.of.birth))/365)

```

## Missing data

Report missing date data (**Date.of.birth** and **Data.collection.date**) and subsequently
unable to calculate **Age.in.years**.

List patients with missing data.

Summarize patients with missing data by recorded **Form.Grade**.

``` {r}

missing.dob <- sum(is.na(as.Date(RawPatientData$Date.of.birth)))
missing.dateofcollection <- sum(is.na(as.Date(RawPatientData$Data.collection.date)))
missing.Age <- sum(is.na(RawPatientData$Age.in.years))

sprintf('Patients with missing Date of Birth: %d', missing.dob)
sprintf('Patients with missing date of data collection : %d', missing.dateofcollection)
sprintf('Patients unable to calculate age: %d', missing.Age)

RawPatientData %>%
  filter(is.na(as.Date(Date.of.birth))) %>%
  select(Full.name,Date.of.birth,SchoolName,Form.Grade,Height) %>%
  kable(caption = 'Patients with no recorded Date of Birth')

RawPatientData %>%
  filter(is.na(as.Date(Date.of.birth))) %>%
  select(Full.name,Date.of.birth,SchoolName,Form.Grade,Height) %>% 
  group_by(Form.Grade) %>% 
  summarize(n=n()) %>%
  kable(caption = 'Recorded Form.Grade of patients with no recorded Date of Birth')

```

## Average age and **Form.Grade**

Summarize mean age of patients by **Form.Grade**, for patients who have calculated **Age.in.years**

``` {r}

RawPatientData %>%
  filter(!is.na(as.Date(Date.of.birth))) %>%
  select(Full.name,Date.of.birth,Age.in.years,SchoolName,Form.Grade,Height) %>%
  group_by(Form.Grade) %>%
  summarize(n=n(), MeanAge = mean(Age.in.years, na.rm = TRUE)) %>%
  kable(caption = 'Mean.Age of patients with calculated age, by Form.Grade')

```


## Dubious data

List and remove

``` {r}

RawPatientData %>%
  filter(as.integer(Age.in.years) <= 0) %>%
  select(Full.name,Date.of.birth,Age.in.years,Gender,Data.collection.date,Form.Grade,Height) %>%
  kable(caption = 'Dubious dates and ages')

# remove dubious data

FilteredAgePatientData <- RawPatientData %>%
  filter(as.integer(Age.in.years) > 0)

```

# Children demographics

``` {r}

# create summary of patients by age and gender (only those aged 18 or below)
p <- FilteredAgePatientData %>%
  filter(as.integer(Age.in.years)<=18) %>%
  mutate(Age = as.integer(Age.in.years)) %>%
  group_by(Age, Gender) %>%
  summarize(n = n())

TotalPatients <- sum(p$n)

MedianAge <- with(FilteredAgePatientData,median(as.numeric(Age.in.years[Age.in.years<=18])))
MeanAge <- with(FilteredAgePatientData,mean(as.numeric(Age.in.years[Age.in.years<=18])))

# ggplot version of plot
gg <- p %>%
  mutate(PlotCount = n*ifelse(Gender == 'Female',1,-1)) %>%
  ggplot(aes(x = Age, y = PlotCount, fill = Gender)) +
  geom_bar(stat = 'identity') + 
  labs(title = "Number of patients in each age group") +
  labs(x = "Age", y = "Number") +
  scale_y_discrete(labels=abs) + 
                     ### remove negative values for axis for 'negative' part of chart
  guides(fill = guide_legend(title="")) +
  theme(legend.position=c(0.8,0.5)) +
  coord_flip()

# plotly version of plot
gp <- p %>%
  spread(Gender, n) %>% # divide male and female into two different variables
  mutate(MalePlot = -Male) %>% # negative male count for 'left'-side of plot
  plot_ly(x = ~Female, y = ~Age, name = 'Female',
          type = 'bar', orientation = 'h',
          hoverinfo = 'text', 
          text = ~paste('Age: ',Age,'years<br>Number :',Female,'<br>Female')) %>%
  add_trace (x = ~MalePlot, name = 'Male',
             hoverinfo = 'text', 
             text = ~paste('Age: ',Age,'years<br>Number :',Male,'<br>Male')) %>%
  layout (xaxis = list(title='Number of children in each age group', tickmode = 'array',
                       tickvals = c(-60, -40, -20, 0, 20, 40, 60),
                       # for 'negative' number, change to absolute value
                       ticktext = c('60', '40', '20', '0', '20', '40', '60')),
          yaxis = list(title='Age'),
          barmode = 'overlay',
          title = 'Spur Afrika 2018 clinics - age of children seen',
          annotations = list(list(x = 40, y = 3, showarrow = F,
                                 text = ~paste('Total children', TotalPatients,
                                               '<br>Median age: ', sprintf('%.1f',MedianAge),
                                               '<br>Mean age: ', sprintf('%.1f',MeanAge)))))

gp

```

# Clinical Findings

Create a 'flattened' database depending on contents of **Findings**

``` {r}

FilteredAgeFindingsData <- FilteredAgePatientData %>%
  mutate(CombinedFindings = Map(append,Findings,Findings2)) %>%
  # Findings were stored in two separate columns
  mutate(Findings = lapply(CombinedFindings,
                           function(Finding) as.data.frame(Finding, stringsAsFactors = FALSE))) %>%
  # unnest requires dataframes
  unnest(Findings) %>% 
  # each finding has its own row, so an individual patient will be listed multiple times
  # change to descriptive finding names
  mutate(Finding = 
           RawFindingCodes$Description[match(Finding,as.character(RawFindingCodes$id))])

```

List the most common findings

``` {r}

FilteredAgeFindingsData %>%
  group_by(Finding) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  kable(caption = 'Findings ranked by frequency')

```

## Dental Findings

``` {r}

FilteredAgeDentalFindingsData <- FilteredAgePatientData %>%
  mutate(Dental.Finding = lapply(Dental.Finding,
                           function(Dental.Find) as.data.frame(Dental.Find, stringsAsFactors = FALSE))) %>%
  # unnest requires dataframes
  unnest(Dental.Finding) %>% 
  # each finding has its own row, so an individual patient will be listed multiple times
  # change to descriptive finding names
  mutate(Dental.Find = 
           RawFindingCodes$Description[match(Dental.Find,as.character(RawFindingCodes$id))])

```

List the most common findings

``` {r}

FilteredAgeDentalFindingsData %>%
  group_by(Dental.Find) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  kable(caption = 'Dental Findings ranked by frequency')
```

# Management

``` {r}

FilteredAgeManagementData <- FilteredAgePatientData %>%
  mutate(Management = lapply(Management,
                           function(Manage) as.data.frame(Manage, stringsAsFactors = FALSE))) %>%
  # unnest requires dataframes
  unnest(Management) %>% 
  # each finding has its own row, so an individual patient will be listed multiple times
  # change to descriptive finding names
  mutate(Manage = 
           RawManagementCodes$Description[match(Manage,as.character(RawManagementCodes$id))])

```

List Management by frequency

``` {r}

FilteredAgeManagementData %>%
  group_by(Manage) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  kable(caption = 'Management ranked by frequency')

```

``` {r}

FilteredAgeManagementDentalData <- FilteredAgePatientData %>%
  mutate(DentalMx = lapply(DentalMx,
                           function(DentalManage) as.data.frame(DentalManage, stringsAsFactors = FALSE))) %>%
  # unnest requires dataframes
  unnest(DentalMx) %>% 
  # each finding has its own row, so an individual patient will be listed multiple times
  # change to descriptive finding names
  mutate(DentalManage = 
           RawManagementCodes$Description[match(DentalManage,as.character(RawManagementCodes$id))])

```

## Dental management

List Dental Management by frequency

``` {r}

FilteredAgeManagementDentalData %>%
  group_by(DentalManage) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  kable(caption = 'Management ranked by frequency')

```

